### 2021-09-14 paper/localpsf.tex
paper cleanup

% REQUIRED
\begin{AMS}
  68Q25, 68R10, 68U05
\end{AMS}

@@

\item[1) Matrix-free:] One cannot easily evaluate kernel entries $\Phi(y,x)$. Instead, access to $\Aop$ is available through ``operator actions.'' That is, we have a black-box computational procedure (which is typically expensive) through which we may evaluate the maps

@@

	u \mapsto\Aop u \quad \text{and} \quad v \mapsto\Aop^Tv

@@

\item[2) Local support:] Let $\phi_x$ be the following \emph{impulse response}:

@@

\phi_x(y) := \Phi(y, x).

@@

We say that $\Aop$ has local support if, for all $x\in \Omega$, the support of $\phi_x$ is contained (or approximately contained) in a neighborhood of $x$. The smaller these supports are, the better our algorithm will perform.  Let $\delta_x$ denote the delta distribution (point source) centered at $x$. The function $\phi_x$ is called the impulse response because straightforward analysis shows that it may be expressed as follows:

@@

\phi_x = \left( \Aop \delta_x^* \right)^*,

@@

where $\Aop \delta_x^* \in L^2(\Omega)'$ is an abuse of notation that denotes the result of applying the operator $\Aop$ to the distribution $\delta_x$, and $\left( \Aop \delta_x^* \right)^* \in L^2(\Omega)$ denotes the Riesz representation of $\Aop \delta_x^*$ with respect to the $L^2$ inner product.\footnote{Recall that the Riesz representative of a functional $\psi \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\psi^* \in L^2(\Omega)$ such that $\psi(v) = \left(\psi^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} Note that while the domain of $\Aop$ is defined as $L^2(\Omega)$, which does not contain $\delta_x$, if $\Phi$ is sufficiently regular then the action of $\Aop$ may be extended to distributions. This is described in Appendix \ref{app:distributions}. From \eqref{eq:impulse_response_delta_action}, we see that this local support property means that the response of $\Aop$ to a point source at $x$ is zero (or small) at points $y$ that are far from $x$.

@@

	\Phi(y+h, x+h) \approx \Phi(y,x)

@@

when $h$ is not too large. One can imagine the kernel entry $\Phi(y,x)$ as representing the strength of an interaction between a source at point $x$ and a target at point $y$. Local translation invariance means that the interaction strength remains roughly the same if the source and target points are both translated by the same vector. This is illustrated in Figure FIG.

@@

\Phi(y,x) \ge 0.

@@

Our motivation for this work is approximation of Hessians in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field, $m$, from noisy observations, 
\begin{equation*}
y=f(m,u(m)) + \text{noise},
\end{equation*}
which depend on  a state variable $u$. In turn, $u$ depends on $m$ implicitly through the solution of a PDE, which we write generically as follows:
\begin{equation}
\label{eq:state_pde}
0=g(m,u).
\end{equation}
Here $u(m)$ denotes the solution of the PDE \eqref{eq:state_pde} as a function of $m$. In the deterministic approach to inverse problems, one typically finds $m$ as the solution to the minimization problem 
\begin{equation}
\label{eq:minimization_problem}
\min_m \quad \frac{1}{2}\|y - f(m,u(m))\|_W^2 + R(m),
\end{equation}
where $\|\cdot\|_W$ is weighted norm which depends on the noise covariance, and $R(m)$ is a regularization term. The objective function in optimization problem \eqref{eq:minimization_problem} is defined indirectly via the implicit function theorem. As a result, the Hessian of the objective function, $\mathcal{H}$, is only accessible matrix-free. In particular, applying $\mathcal{H}$ to a vector requires solving two PDEs \cite{HESSIANACTION}. More accessible approximations of $\mathcal{H}$ are highly desirable, because they allow for fast solution of \eqref{eq:minimization_problem} via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because $\mathcal{H}^{-1}$ locally approximates the Bayesian posterior covariance for $m$.

@@

If $\Aop$ were perfectly translation invariant (i.e., if equality held in \eqref{eq:local_translation_invariance} for all $x$, $y$, $h$) then it is straightforward to show that $\Aop$ would be the convolution operator $u \mapsto \left(\varphi_x \ast u\right)^*$, where the convolution kernel is given by
\begin{equation}
\label{eq:convolution_kernel}
	\varphi_x(y) := \phi_x(y+x).
\end{equation}
The convolution kernel $\varphi_x$ is the result of translating the impulse response $\phi_x$ to re-center it at zero instead of $x$. Locally translation invariant operators act like convolution operators locally, but the convolution kernel, $\varphi_x$, varies as $x$ changes. We therefore approximate $\Aop$ by a spatially varying weighted sum of convolution operators, $\AopPc \approx \Aop$, of the form
\begin{equation}
\label{eq:product_convolution}
	\left(\AopPc~u\right)(v) := \left( v,~ \sum_{i=1}^r \varphi_i \ast \left(w_i \cdot u\right) \right)_{L^2(\Omega)},
\end{equation}
where $f \cdot g$ denotes pointwise multiplication, $f \ast g$ denotes convolution, and $\left(f, g\right)_{L^2(\Omega)}$ denotes the $L^2$ inner product on $\Omega$, for functions $f,g \in L^2(\Omega)$. Approximations of the form \eqref{eq:product_convolution} are known as \emph{product-convolution} approximations, because the action of each term in the sum consists of a pointwise product, followed by a convolution. To avoid nested subscripts, we write 
\begin{equation*}
\varphi_i := \varphi_{x_i} \quad \text{and} \quad \phi_i := \phi_{x_i},
\end{equation*}
with $\varphi_{x_i}$ defined in \eqref{eq:convolution_kernel} and $\phi_{x_i}$ defined in \eqref{eq:impulse_response_defn}, to denote local convolution kernels and impulse responses corresponding to a collection of points $\{x_i\}_{i=1}^r \subset \Omega$ scattered throughout the domain. The functions $w_i$ are spatially varying weighting functions that are used to interpolate the convolution kernels $\varphi_i$. Our operator approximation method is defined by how we compute the convolution kernels, $\varphi_i$, how we choose the points, $x_i$, and what weighting functions, $w_i$, that we use. In Section \ref{sec:overview_intro} we summarize the answers to these questions; we will provide detailed answers to these questions in Section \ref{sec:method}.

@@

We compute one ``batch'' of $\varphi_i$'s by applying $\Aop$ to a sum of point sources (Dirac comb) associated with a collection of points $x_i$ scattered throughout $\Omega$. The batch of points $x_i$ are chosen so that the support of $\phi_i$ and the support of $\phi_j$ do not overlap (or do not overlap much) if $i \neq j$. Because these supports do not overlap, we can post-process the response of $\Aop$ to the Dirac comb to recover the functions $\phi_i$, and therefore the functions $\varphi_i$, associated with all points $x_i$ in the batch---with one application of $\Aop$, we recover many $\varphi_i$. This is illustrated in Figure FIG. The process is repeated to get more batches of $\varphi_i$'s, until a desired number of batches is reached.

In order to choose the points $x_i$, we need to estimate the support of $\phi_i$ \emph{before} we compute it. The supports of the functions $\phi_x$ are estimated a-priori, for all $x \in \Omega$ simultaneously, via a procedure that involves applying $\Aop^T$ to a small number of specially chosen functions (three functions if $d=1$, six if $d=2$, ten if $d=3$), then post processing the results (Section \ref{eq:mean_and_covariance_estimation}). This procedure is the only part of the paper that requires non-negativity of the integral kernel.

@@

Once the $\varphi_i$ and $w_i$ are computed, we convert $\AopPc$ to hierarchical matrix (H-matrix) format. Let $\Phi_\text{pc}$ denote the integral kernel associated with $\AopPc$. While $\Phi(y,x)$ is not easily computable, $\Phi_\text{pc}(y,x)$ is given by the formula
\begin{equation}
\label{eq:kernel_entries}
	\Phi_\text{pc}(y,x) = \sum_{i=1}^r \varphi_i(y-x) w_i(x).
\end{equation}
This follows from~\eqref{eq:product_convolution} and the fact that the convolution operator $u \mapsto \varphi_i \ast u$ has $(y,x)$ kernel entry given by $\varphi(y-x)$.
Formula~\eqref{eq:kernel_entries} allows us to construct a H-matrix representation of $\AopPc$ using the conventional adaptive cross H-matrix construction method, in which one forms low-rank approximations of blocks of the matrix by sampling rows and columns of those blocks \cite{HACA}. Once in H-matrix format, fast H-matrix arithmetic is used to invert $\AopPc$, or perform other useful matrix operations.

@@

We choose sample points, $x_i$, in batches. A greedy algorithm is used to choose as many points as possible per batch, while ensuring that the points in each batch are not too close to each other. We ensure that the points $x_i$ are well separated from each other by forming a-priori estimates of the supports of the functions $\phi_x$ (Section \ref{eq:mean_and_covariance_estimation}). The support of $\phi_x$ is estimated to be contained within an ellipsoid, so the process of choosing batches of points is an ellipsoid packing problem (Section \ref{sec:greedy_point_selection}).

@@

Because of the non-negative kernel property, for each $x \in \Omega$, the impulse response $\phi_x$ is a non-negative function $y \mapsto \phi_x(y)$. Hence, $\phi_x$ is a scaled probability distribution. Let $\alpha$ denote the scaling factor, let $\widehat{\phi}_x := \phi_x / \alpha(x)$ denote the normalized version of $\phi_x$, and let $\mu(x)$ and $\Sigma(x)$ denote the mean and covariance of $\widehat{\phi}_x$, respectively. We make the approximation that the support of $\phi_x$ is contained within the ellipsoid
\begin{equation}
\label{eq:support_ellipsoid}
E_x := \{x' \in \Omega: (x' - \mu(x))^T \Sigma(x)^{-1} (x' - \mu(x)) \le \tau^2\},
\end{equation}
where $\tau$ is a fixed constant. The ellipsoid $E_x$ is the set of points within $\tau$ standard deviations from the mean of the Gaussian distribution with mean $\mu(x)$ and covariance $\Sigma(x)$, i.e., the Gaussian distribution which has the same mean and covariance as $\widehat{\phi}_x$ (see Figure FIG). The quantity $\tau$ is a parameter that must be chosen by the user of the algorithm. The larger $\tau$ is, the larger the ellipsoid $E_x$ is, and the more conservative the estimate is for the support of $\phi_x$. However, the cost of our algorithm will depend on how many non-overlapping ellipsoids $E_x$ we can ``pack'' in the domain $\Omega$ (more ellipsoids $\implies$ lower cost), and choosing a larger value of $\tau$ means that fewer ellipsoids will fit in $\Omega$. We find that $\tau=3$ yields a reasonable balance between these competing interests, and use $\tau=3$ in our numerical results. The fraction of the ``mass'' of $\phi_x$ residing outside of $E_x$ is less than $1/\tau^2$ by Chebyshev's inequality, though this bound is conservative and typically far less mass resides in this region.

The ellipsoid $E_x$ is not immediately available. To use $E_x$, we must first compute the scalar $\alpha(x)$, the length-$d$ vector $\mu(x)$, and the $d \times d$ matrix $\Sigma(x)$ (recall $d \in \{1,2,3\}$ is the spatial dimension of $\Omega$). The direct approach to compute $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$ is to apply $\Aop$ to a point source centered at $x$ to get $\phi_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\phi_x$ to determine $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$. But this direct approach is computationally infeasible because and it requires one operator action of $\Aop$ per point $x$ that we wish to evaluate $E_x$ at, and we will need to evaluate $E_x$ at a large number of points. Fortunately, it is possible to compute $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$, \emph{for all points $x$ simultaneously}, by applying $\Aop^T$ to one constant function, $d$ linear functions, and $d(d+1)/2$ quadratic functions. We present this efficient method of computing $\alpha$, $\mu$, and $\Sigma$ in Theorem \ref{thm:vol_mean_cov} and summarize the method in Algorithm \ref{alg:varhpi_mean_cov}.

\begin{theorem}[Impulse response moments]
	\label{thm:vol_mean_cov}
	Let $\alpha:\Omega \rightarrow \mathbb{R}$, $\mu:\Omega \rightarrow \mathbb{R}^d$, and $\Sigma:\Omega \rightarrow \mathbb{R}^{d \times d}$, be the following functions:
	\begin{equation*}
	\alpha(x) := \int_{\Omega} \phi_x(y) dy, \qquad
	\mu(x) := \mathbb{E}(\widehat{\phi}_x), \qquad
	\Sigma(x) := \Var(\widehat{\phi}_x),
	\end{equation*}
	where $\mathbb{E}(\widehat{\phi}_x)$ denotes the mean of $\widehat{\phi}_x$, and $\Var(\widehat{\phi}_x)$ denotes the variance of $\widehat{\phi}_x$.	Let  $x^i$ denote the $i^\text{th}$ component of $x$, i.e., $x = \left(x^1, x^2, \dots, x^d\right)$. Let $C$, $\{L^i\}_{i=1}^d$, and ${\{Q^{ij}\}_{i=1}^d}_{j=1}^d$ be the following constant, linear, and quadratic functions:
	\begin{equation*}
	C(x) := 1, \qquad
	L^i(x) := x^i, \qquad
	Q^{ij}(x) = x^i x^j.
	\end{equation*}
	Let $f/g$ to denote the pointwise division of functions, $\left(f/g\right)(x) = f(x)/g(x)$, and $f\cdot g$ to denote the pointwise multiplication of functions, $(f\cdot g)(x) = f(x)g(x)$. The following results hold:
	\begin{subequations}
		\label{eq:vol_mean_var}
	\begin{align}
	\alpha =& \left(\Aop^T C\right)^* \\
	\mu^i =& \left(\Aop^T L^i\right)^* / \alpha \\
	\Sigma^{ij} =& \left(\Aop^T Q^{ij}\right)^* / \alpha - \mu^i\cdot \mu^j
	\end{align}
	\end{subequations}
	 for $i=1,\dots, d$, $j=1,\dots,d$.
\end{theorem}

@@

We use a greedy ellipsoid packing algorithm to choose sets of sample points, $S_b$, such that there is no overlap between the ellipsoids, $E_{x_i}$, associated with the sample points, $x_i$, within a batch. These sample point batches will be used in Section \ref{sec:get_impulse_response} to compute many impulse responses of $\Aop$ using only a small number of operator actions of $\Aop$. The support of $\phi_x$ is (approximately) contained in the ellipsoid $E_x$, so by applying $\mathcal{A}$ to the Dirac comb associated with all sample points in a batch, we will recover the impulse responses for all sample points in that batch.

@@

We compute convolution kernels $\varphi_{x_i}$ in batches by applying $\Aop$ to a small number Dirac combs. Each application of $\Aop$ to a Dirac comb yields one batch of impulse responses, $\phi_{x_i}$, which yields one batch of convolution kernels, $\varphi_{x_i}$. The Dirac comb, $\dcomb_b$, associated with a batch of sample points, $S_b$, is the sum of Dirac distributions (point sources) centered at the points $x_i \in S_b$. That is,
\begin{equation*}
	\dcomb_b := \sum_{x_i \in S_b} \delta_{x_i}.
\end{equation*}
For each batch $S_b$, we compute the action of $\mathcal{A}$ on the associated Dirac comb:
\begin{equation}
\label{eq:dirac_comb_H_action}
	\combresponse_b := \left(\mathcal{A} \dcomb_b^*\right)^*.
\end{equation}
By linearity, the resulting function $\combresponse_b$ may be written as
\begin{equation}
\label{eq:phi_b}
	\combresponse_b = \left\langle\mathcal{A},\sum_{x_i \in S_b} \delta_{x_i}\right\rangle^* = \sum_{x_i \in S_b} \langle\mathcal{A}, \delta_{x_i}\rangle^* = \sum_{x_i \in S_b} \phi_{x_i}.
\end{equation}
Since the points $x_i$ are chosen so that the ellipsoid $E_{x_i}$ that (approximately) supports $\phi_i$, and the ellipsoid $E_{x_j}$ that (approximately) supports $\phi_j$ do not overlap when $i \neq j$, from \eqref{eq:convolution_kernel} and \eqref{eq:phi_b} we have (approximately)
\begin{equation}
\label{eq:varphi_eval}
	\varphi_{x_i}(z) = \phi_i(z+x_i) = \begin{cases}
		\combresponse_b(z+x_i), & z+x_i \in E_{x_i} \\
		0, & \text{otherwise}
		\end{cases}
\end{equation}
for all $x_i \in S_b$. By performing one matrix-vector product, \eqref{eq:dirac_comb_H_action}, we recover the convolution kernels $\varphi_{x_i}$ associated with every point $x_i \in S_b$.

@@

%		\tcp{Compute weighting functions $\{w_i\}_{i=1}^r$}

@@

%We summarize coefficient vector versions of the function space operation that are used in this paper in Appendix \ref{app:discretized_operations}.
%
%As described in Section \ref{sec:discretization}, a function $u_h$ in a finite element space $V_h$ may be written in terms of its vector of coefficients, $\mathbf{u} \in \mathbb{R}^N_\mathbf{M}$, with respect the finite element basis $\{\phi_i\}_{i=1}^N$. Recall that $\mathbf{M}$ is the mass matrix. We have
%\begin{equation*}
%	u(x) \approx u_h(x) = \sum_{i=1}^N \mathbf{u}_i \phi_i(x).
%\end{equation*}
%The finite element function space $V_h$ is isometrically isomorphic to the coefficient space $\mathbb{R}^N_\mathbf{M}$, and the dual space $V_h'$ is isometrically isomorphic to $\mathbb{R}^N_{\mathbf{M}^{-1}}$. In computations, we perform operations with coefficient vectors in $\mathbb{R}^N_\mathbf{M}$ and coefficient dual vectors in $\mathbb{R}^N_{\mathbf{M}^{-1}}$, rather than functions in $V_h$ and functionals in $V_h'$. Here we describe how to perform common function space operations in terms of coefficient vectors and coefficient dual vectors.

>>>>

% REQUIRED
\begin{AMS}
  ASDF
\end{AMS}

@@

\item[1) Matrix-free:] One cannot easily evaluate kernel entries $\Aker(y,x)$. Instead, access to $\Aop$ is available through ``operator actions.'' That is, we have a black-box computational procedure (which is typically expensive) through which we may evaluate the maps

@@

	u \mapsto\Aop u \quad \text{and} \quad v \mapsto \Aop^Tv

@@

\item[2) Local support:] Let $\impulseresponse_x$ be the following \emph{impulse response}:

@@

\impulseresponse_x(y) := \Aker(y, x).

@@

We say that $\Aop$ has local support if, for all $x\in \Omega$, the support of $\impulseresponse_x$ is contained (or approximately contained) in a neighborhood of $x$. The smaller these supports are, the better our algorithm will perform.  Let $\delta_x$ denote the delta distribution (point source) centered at $x$. The function $\impulseresponse_x$ is called the impulse response because straightforward analysis shows that it may be expressed as follows:

@@

	\impulseresponse_x = \left( \Aop \delta_x^* \right)^*,

@@

where $\Aop \delta_x^* \in L^2(\Omega)'$ is an abuse of notation that denotes the result of applying the operator $\Aop$ to the distribution $\delta_x$, and $\left( \Aop \delta_x^* \right)^* \in L^2(\Omega)$ denotes the Riesz representation of $\Aop \delta_x^*$ with respect to the $L^2$ inner product.\footnote{Recall that the Riesz representative of a functional $\psi \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\psi^* \in L^2(\Omega)$ such that $\psi(v) = \left(\psi^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} Note that while the domain of $\Aop$ is defined as $L^2(\Omega)$, which does not contain $\delta_x$, if $\Aker$ is sufficiently regular then the action of $\Aop$ may be extended to distributions. This is described in Appendix \ref{app:distributions}. From \eqref{eq:impulse_response_delta_action}, we see that this local support property means that the response of $\Aop$ to a point source at $x$ is zero (or small) at points $y$ that are far from $x$.

@@

	\Aker(y+h, x+h) \approx \Aker(y,x)

@@

when $h$ is not too large. One can imagine the kernel entry $\Aker(y,x)$ as representing the strength of an interaction between a source at point $x$ and a target at point $y$. Local translation invariance means that the interaction strength remains roughly the same if the source and target points are both translated by the same vector. This is illustrated in Figure FIG.

@@

\Aker(y,x) \ge 0.

@@

Our motivation for this work is approximation of Hessians in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field, $m$, from noisy observations, 
\begin{equation*}
	y = f(m, u(m)) + \text{noise},
\end{equation*}
which depend on  a state variable $u$. In turn, $u$ depends on $m$ implicitly through the solution of a PDE, which we write generically as follows:
\begin{equation}
	\label{eq:state_pde}
	0 = g(m,u).
\end{equation}
Here $u(m)$ denotes the solution of the PDE \eqref{eq:state_pde} as a function of $m$. In the deterministic approach to inverse problems, one typically finds $m$ as the solution to the minimization problem 
\begin{equation}
	\label{eq:minimization_problem}
	\min_m \quad \frac{1}{2}\|y - f(m,u(m))\|_W^2 + R(m),
\end{equation}
where $\|\cdot\|_W$ is weighted norm which depends on the noise covariance, and $R(m)$ is a regularization term. The objective function in optimization problem \eqref{eq:minimization_problem} is defined indirectly via the implicit function theorem. As a result, the Hessian of the objective function, $\mathcal{H}$, is only accessible matrix-free. In particular, applying $\mathcal{H}$ to a vector requires solving two PDEs \cite{HESSIANACTION}. More accessible approximations of $\mathcal{H}$ are highly desirable, because they allow for fast solution of \eqref{eq:minimization_problem} via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because $\mathcal{H}^{-1}$ locally approximates the Bayesian posterior covariance for $m$.

@@

If $\Aop$ were perfectly translation invariant (i.e., if equality held in \eqref{eq:local_translation_invariance} for all $x$, $y$, $h$) then it is straightforward to show that $\Aop$ would be the convolution operator $u \mapsto \left(\varphi_x \ast u\right)^*$, where the convolution kernel is given by
\begin{equation}
	\label{eq:convolution_kernel}
	\varphi_x(y) := \impulseresponse_x(y+x).
\end{equation}
The convolution kernel $\varphi_x$ is the result of translating the impulse response $\impulseresponse_x$ to re-center it at zero instead of $x$. Locally translation invariant operators act like convolution operators locally, but the convolution kernel, $\varphi_x$, varies as $x$ changes. We therefore approximate $\Aop$ by a spatially varying weighted sum of convolution operators, $\AopPc \approx \Aop$, of the form
\begin{equation}
\label{eq:product_convolution}
	\left(\AopPc~u\right)(v) := \left( v,~ \sum_{i=1}^r \varphi_i \ast \left(w_i \cdot u\right) \right)_{L^2(\Omega)},
\end{equation}
where $f \cdot g$ denotes pointwise multiplication, $f \ast g$ denotes convolution, and $\left(f, g\right)_{L^2(\Omega)}$ denotes the $L^2$ inner product on $\Omega$, for functions $f,g \in L^2(\Omega)$. Approximations of the form \eqref{eq:product_convolution} are known as \emph{product-convolution} approximations, because the action of each term in the sum consists of a pointwise product, followed by a convolution. To avoid nested subscripts, we write 
\begin{equation*}
	\varphi_i := \varphi_{x_i} \quad \text{and} \quad \impulseresponse_i := \impulseresponse_{x_i},
\end{equation*}
with $\varphi_{x_i}$ defined in \eqref{eq:convolution_kernel} and $\impulseresponse_{x_i}$ defined in \eqref{eq:impulse_response_defn}, to denote local convolution kernels and impulse responses corresponding to a collection of points $\{x_i\}_{i=1}^r \subset \Omega$ scattered throughout the domain. The functions $w_i$ are spatially varying weighting functions that are used to interpolate the convolution kernels $\varphi_i$. Our operator approximation method is defined by how we compute the convolution kernels, $\varphi_i$, how we choose the points, $x_i$, and what weighting functions, $w_i$, that we use. In Section \ref{sec:overview_intro} we summarize the answers to these questions; we will provide detailed answers to these questions in Section \ref{sec:method}.

@@

We compute one ``batch'' of $\varphi_i$'s by applying $\Aop$ to a sum of point sources (Dirac comb) associated with a collection of points $x_i$ scattered throughout $\Omega$. The batch of points $x_i$ are chosen so that the support of $\impulseresponse_i$ and the support of $\impulseresponse_j$ do not overlap (or do not overlap much) if $i \neq j$. Because these supports do not overlap, we can post-process the response of $\Aop$ to the Dirac comb to recover the functions $\impulseresponse_i$, and therefore the functions $\varphi_i$, associated with all points $x_i$ in the batch---with one application of $\Aop$, we recover many $\varphi_i$. This is illustrated in Figure FIG. The process is repeated to get more batches of $\varphi_i$'s, until a desired number of batches is reached.

In order to choose the points $x_i$, we need to estimate the support of $\impulseresponse_i$ \emph{before} we compute it. The supports of the functions $\impulseresponse_x$ are estimated a-priori, for all $x \in \Omega$ simultaneously, via a procedure that involves applying $\Aop^T$ to a small number of specially chosen functions (three functions if $d=1$, six if $d=2$, ten if $d=3$), then post processing the results (Section \ref{eq:mean_and_covariance_estimation}). This procedure is the only part of the paper that requires non-negativity of the integral kernel.

@@

Once the $\varphi_i$ and $w_i$ are computed, we convert $\AopPc$ to hierarchical matrix (H-matrix) format. Let $\Aker_\text{pc}$ denote the integral kernel associated with $\AopPc$. While $\Aker(y,x)$ is not easily computable, $\Aker_\text{pc}(y,x)$ is given by the formula
\begin{equation}
	\label{eq:kernel_entries}
	\AkerPc(y,x) = \sum_{i=1}^r \varphi_i(y-x) w_i(x).
\end{equation}
This follows from~\eqref{eq:product_convolution} and the fact that the convolution operator $u \mapsto \varphi_i \ast u$ has $(y,x)$ kernel entry given by $\varphi(y-x)$.
Formula~\eqref{eq:kernel_entries} allows us to construct a H-matrix representation of $\AopPc$ using the conventional adaptive cross H-matrix construction method, in which one forms low-rank approximations of blocks of the matrix by sampling rows and columns of those blocks \cite{HACA}. Once in H-matrix format, fast H-matrix arithmetic is used to invert $\AopPc$, or perform other useful matrix operations.

@@

We choose sample points, $x_i$, in batches. A greedy algorithm is used to choose as many points as possible per batch, while ensuring that the points in each batch are not too close to each other. We ensure that the points $x_i$ are well separated from each other by forming a-priori estimates of the supports of the functions $\impulseresponse_x$ (Section \ref{eq:mean_and_covariance_estimation}). The support of $\impulseresponse_x$ is estimated to be contained within an ellipsoid, so the process of choosing batches of points is an ellipsoid packing problem (Section \ref{sec:greedy_point_selection}).

@@

Because of the non-negative kernel property, for each $x \in \Omega$, the impulse response $\impulseresponse_x$ is a non-negative function $y \mapsto \impulseresponse_x(y)$. Hence, $\impulseresponse_x$ is a scaled probability distribution. Let $\alpha$ denote the scaling factor, let $\widehat{\impulseresponse}_x := \impulseresponse_x / \alpha(x)$ denote the normalized version of $\impulseresponse_x$, and let $\mu(x)$ and $\Sigma(x)$ denote the mean and covariance of $\widehat{\impulseresponse}_x$, respectively. We make the approximation that the support of $\impulseresponse_x$ is contained within the ellipsoid
\begin{equation}
	\label{eq:support_ellipsoid}
	E_x := \{x' \in \Omega: (x' - \mu(x))^T \Sigma(x)^{-1} (x' - \mu(x)) \le \tau^2\},
\end{equation}
where $\tau$ is a fixed constant. The ellipsoid $E_x$ is the set of points within $\tau$ standard deviations from the mean of the Gaussian distribution with mean $\mu(x)$ and covariance $\Sigma(x)$, i.e., the Gaussian distribution which has the same mean and covariance as $\widehat{\impulseresponse}_x$ (see Figure FIG). The quantity $\tau$ is a parameter that must be chosen by the user of the algorithm. The larger $\tau$ is, the larger the ellipsoid $E_x$ is, and the more conservative the estimate is for the support of $\impulseresponse_x$. However, the cost of our algorithm will depend on how many non-overlapping ellipsoids $E_x$ we can ``pack'' in the domain $\Omega$ (more ellipsoids $\implies$ lower cost), and choosing a larger value of $\tau$ means that fewer ellipsoids will fit in $\Omega$. We find that $\tau=3$ yields a reasonable balance between these competing interests, and use $\tau=3$ in our numerical results. The fraction of the ``mass'' of $\impulseresponse_x$ residing outside of $E_x$ is less than $1/\tau^2$ by Chebyshev's inequality, though this bound is conservative and typically far less mass resides in this region.

The ellipsoid $E_x$ is not immediately available. To use $E_x$, we must first compute the scalar $\alpha(x)$, the length-$d$ vector $\mu(x)$, and the $d \times d$ matrix $\Sigma(x)$ (recall $d \in \{1,2,3\}$ is the spatial dimension of $\Omega$). The direct approach to compute $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$ is to apply $\Aop$ to a point source centered at $x$ to get $\impulseresponse_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\impulseresponse_x$ to determine $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$. But this direct approach is computationally infeasible because and it requires one operator action of $\Aop$ per point $x$ that we wish to evaluate $E_x$ at, and we will need to evaluate $E_x$ at a large number of points. Fortunately, it is possible to compute $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$, \emph{for all points $x$ simultaneously}, by applying $\Aop^T$ to one constant function, $d$ linear functions, and $d(d+1)/2$ quadratic functions. We present this efficient method of computing $\alpha$, $\mu$, and $\Sigma$ in Theorem \ref{thm:vol_mean_cov} and summarize the method in Algorithm \ref{alg:varhpi_mean_cov}.

\begin{theorem}[Impulse response moments]
	\label{thm:vol_mean_cov}
	Let $\alpha:\Omega \rightarrow \mathbb{R}$, $\mu:\Omega \rightarrow \mathbb{R}^d$, and $\Sigma:\Omega \rightarrow \mathbb{R}^{d \times d}$, be the following functions:
	\begin{equation*}
		\alpha(x) := \int_{\Omega} \impulseresponse_x(y) dy, \qquad
		\mu(x) := \mathbb{E}(\widehat{\impulseresponse}_x), \qquad
		\Sigma(x) := \Var(\widehat{\impulseresponse}_x),
	\end{equation*}
	where $\mathbb{E}(\widehat{\impulseresponse}_x)$ denotes the mean of $\widehat{\impulseresponse}_x$, and $\Var(\widehat{\impulseresponse}_x)$ denotes the variance of $\widehat{\impulseresponse}_x$.	Let  $x^i$ denote the $i^\text{th}$ component of $x$, i.e., $x = \left(x^1, x^2, \dots, x^d\right)$. Let $C$, $\{L^i\}_{i=1}^d$, and ${\{Q^{ij}\}_{i=1}^d}_{j=1}^d$ be the following constant, linear, and quadratic functions:
	\begin{equation*}
		C(x) := 1, \qquad
		L^i(x) := x^i, \qquad
		Q^{ij}(x) = x^i x^j.
	\end{equation*}
	Let $f/g$ to denote the pointwise division of functions, $\left(f/g\right)(x) = f(x)/g(x)$, and $f\cdot g$ to denote the pointwise multiplication of functions, $(f\cdot g)(x) = f(x)g(x)$. The following results hold:
	\begin{subequations}
		\label{eq:vol_mean_var}
		\begin{align}
			\alpha =& \left(\Aop^T C\right)^* \\
			\mu^i =& \left(\Aop^T L^i\right)^* / \alpha \\
			\Sigma^{ij} =& \left(\Aop^T Q^{ij}\right)^* / \alpha - \mu^i\cdot \mu^j
		\end{align}
	\end{subequations}
	 for $i=1,\dots, d$, $j=1,\dots,d$.
\end{theorem}

@@

We use a greedy ellipsoid packing algorithm to choose sets of sample points, $S_b$, such that there is no overlap between the ellipsoids, $E_{x_i}$, associated with the sample points, $x_i$, within a batch. These sample point batches will be used in Section \ref{sec:get_impulse_response} to compute many impulse responses of $\Aop$ using only a small number of operator actions of $\Aop$. The support of $\impulseresponse_x$ is (approximately) contained in the ellipsoid $E_x$, so by applying $\mathcal{A}$ to the Dirac comb associated with all sample points in a batch, we will recover the impulse responses for all sample points in that batch.

@@

We compute convolution kernels $\varphi_{x_i}$ in batches by applying $\Aop$ to a small number Dirac combs. Each application of $\Aop$ to a Dirac comb yields one batch of impulse responses, $\impulseresponse_{x_i}$, which yields one batch of convolution kernels, $\varphi_{x_i}$. The Dirac comb, $\dcomb_b$, associated with a batch of sample points, $S_b$, is the sum of Dirac distributions (point sources) centered at the points $x_i \in S_b$. That is,
\begin{equation*}
	\dcomb_b := \sum_{x_i \in S_b} \delta_{x_i}.
\end{equation*}
For each batch $S_b$, we compute the action of $\mathcal{A}$ on the associated Dirac comb:
\begin{equation}
	\label{eq:dirac_comb_H_action}
	\combresponse_b := \left(\mathcal{A} \dcomb_b^*\right)^*.
\end{equation}
By linearity, the resulting function $\combresponse_b$ may be written as
\begin{equation}
	\label{eq:phi_b}
	\combresponse_b = \left\langle\mathcal{A},\sum_{x_i \in S_b} \delta_{x_i}\right\rangle^* = \sum_{x_i \in S_b} \langle\mathcal{A}, \delta_{x_i}\rangle^* = \sum_{x_i \in S_b} \impulseresponse_{x_i}.
\end{equation}
Since the points $x_i$ are chosen so that the ellipsoid $E_{x_i}$ that (approximately) supports $\impulseresponse_i$, and the ellipsoid $E_{x_j}$ that (approximately) supports $\impulseresponse_j$ do not overlap when $i \neq j$, from \eqref{eq:convolution_kernel} and \eqref{eq:phi_b} we have (approximately)
\begin{equation}
\label{eq:varphi_eval}
	\varphi_{x_i}(z) = \impulseresponse_i(z+x_i) = 
	\begin{cases}
		\combresponse_b(z+x_i), & z+x_i \in E_{x_i} \\
		0, & \text{otherwise}
	\end{cases}
\end{equation}
for all $x_i \in S_b$. By performing one matrix-vector product, \eqref{eq:dirac_comb_H_action}, we recover the convolution kernels $\varphi_{x_i}$ associated with every point $x_i \in S_b$.

@@

@@
